---
title: "CodeBook"
author: "rch"
date: "Monday, January 12, 2015"
output: html_document
---

#Codebook for Getting and Cleaning Data Course Project
======================================================

##Data
This script is to take the UCI HAR (Human Activity Recognition 
      using smart phones) Dataset and do the following things:
Create one R script called run_analysis.R that does the following. 
1. Merge the training and the test sets to create one data set.
2. Extract only the measurements on the mean and standard deviation for each 
  measurement. 
3. Use descriptive activity names to name the activities in the data set
4. Appropriately label the data set with descriptive variable names. 
5. From the data set in step 4, create a second, independent tidy data set with 
  the average of each variable for each activity and each subject.

##Transformation Process and Variables
###The following markdown documents the processes taken and the code used.

Set the working directory to find the necessary files
```{r}
setwd("~/R_Files/Coursera_GettingCleaningData")
```

Load in the required libraries
```{r}
library(utils)
library(dplyr)
library(stringr)
library(tidyr)
```

###Step 1
Read in training data sets and apply initial column names
```{r}
features <- read.table("./data/UCI HAR Dataset/features.txt", 
                       stringsAsFactors = FALSE)
xtrn <- read.table("./data/UCI HAR Dataset/train/X_train.txt", nrows = 7352)
colnames(xtrn) <- features$V2
ytrn <- read.table("./data/UCI HAR Dataset/train/y_train.txt")
colnames(ytrn) <- "Activity"
strn <- read.table("./data/UCI HAR Dataset/train/subject_train.txt")
colnames(strn) <- "Subject"
train <- cbind(strn, ytrn, xtrn)
```

Read in test data sets and apply initial column names
```{r}
xtest <- read.table("./data/UCI HAR Dataset/test/X_test.txt", nrows = 2947)
colnames(xtest) <- features$V2
ytest <- read.table("./data/UCI HAR Dataset/test/y_test.txt")
colnames(ytest) <- "Activity"
stest <- read.table("./data/UCI HAR Dataset/test/subject_test.txt")
colnames(stest) <- "Subject"
test <- cbind(stest, ytest, xtest)
```

Combine both the train and test data sets into a single data frame
```{r}
df <- rbind(test, train)
```

###Step 2
Filter columns to just the activity, subject, and measurement columns
         with mean and std (standard deviation) and call data frame skinny
```{r}
keepnames <- grepl("mean|std|Subject|Activity", colnames(df)) & 
  !grepl("meanFreq", colnames(df))
skinny <- df[, keepnames]
```

###Step 3
Change Activities from numeric to names
Load in activity names file and add appropriate column names prior to join
```{r}
activity_names <- read.table("./data/UCI HAR Dataset/activity_labels.txt", 
                             stringsAsFactors = FALSE)
colnames(activity_names) <- c("Activity", "ActDesc")
```

Join activity names with skinny by Activity in each data frame
```{r}
skinny <- tbl_df(skinny)
skinny <- inner_join(skinny, activity_names)
```

Reorder columns to move new activity names to first column 
 and remove and rename Description as activity
```{r}
skinny <- skinny[,c(69, 2:68)]
colnames(skinny)[1] <- "Activity"
```

###Step 4

Rename variable names in skinny to make more sense and separate by "-"
```{r}
pracnames <- names(skinny)
```

Change Acc to acceleration
```{r}
pracnames <- str_replace_all(pracnames, "Acc", "-acceleration-")
```

Change Gyro to gyroscope
```{r}
pracnames <- str_replace_all(pracnames, "Gyro", "-gyroscope-")
```

Change Mag to magnitude
```{r}
pracnames <- str_replace_all(pracnames, "Mag", "-magnitude")
```

Remove ()
```{r}
pracnames <- str_replace_all(pracnames, "\\(\\)", "")
```

Change t at the beginning to time
```{r}
pracnames <- str_replace_all(pracnames, "^t", "time-")
```

Change f at the beginning to frequency
```{r}
pracnames <- str_replace_all(pracnames, "^f", "frequency-")
```

Replace tBody with time-body
```{r}
pracnames <- str_replace_all(pracnames, "tBody", "time-body-")
```

Replace BodyBody with body
```{r}
pracnames <- str_replace_all(pracnames, "BodyBody", "body")
```

Clean up by removing any "--" and change to "-"
```{r}
pracnames <- str_replace_all(pracnames, "--", "-")
```

Make all lower case
```{r}
pracnames <- tolower(pracnames)
```

Apply them to skinny data frame
```{r}
names(skinny) <- pracnames
```

###Step 5
Use tidyr to gather measurements into tidy data (1 line per observation)
```{r}
ttidy <- skinny %>%
  gather(sensor, Value, 3:68)
```

Use dplyr to summarize by subject and activity
```{r}
ttidy <- ttidy %>%
  group_by(subject, activity)%>%
  summarize(mean = mean(Value))
```

Write out tidy data set to file
```{r}
write.csv(ttidy, "tidy_data.csv", row.names = FALSE)
```

